A Collection of Thoughts on the Keccak Construction

### 2018-04-10 s2018-math490/thesis/thesis.tex
Updates to thesis; fleshed out bijectivity sections
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

\title{Keccak Thesis}

@@

\section{A Series of Introductions} \label{sec:intro}
    This paper presents the work of the author towards the completion of the
requirements for the MATH 490 Independent Study course under Dr. Clifford
Bergman at Iowa State University of Science and Technology. This work was an
extension of the author's Honors Project under Dr. Eric Bergman and utilized
the resulting
\href{https://github.com/cipherboy/hash_framework}{hash\_framework} project.
Additional artifacts related to this project can be seen in the
\href{https://github.com/cipherboy/keccak-attacks}{keccak-attacks} repository.

@@

    While there have been many applications of mathematics, few are as
demanding and shrouded in secrecy as that of cryptography. Cryptography exists
because of the fundamental need of civilizations, governments, and individuals
to keep secrets secure from devoted adversaries.

@@

@@

@@

\subsection{Bijectivity} \label{sec:p:bijectivity}

% TODO Outline
%   - Reason why bijectivity is important
%   - Impact on attacks?

\subsubsection{Of $\theta$} \label{sec:p:b:t}

@@

\subsubsection{Of $\rho$} \label{sec:p:b:r}

@@

\subsubsection{Of $\pi$} \label{sec:p:b:p}

@@

\subsubsection{Of $\chi$} \label{sec:p:b:c}

@@

\subsubsection{Of $\iota$} \label{sec:p:b:i}

@@

\subsection{Order of the Permutation} \label{sec:p:order}

% TODO Outline
%   - Outline why interesting
%   - Attack on PRF/XOF -> low cycle size
%   - Values over 2^(25w) indicate PRF has smaller upper bound for actual
%       entropy
%   - T R P C I

@@

\subsection{Inverse of the Permutation} \label{sec:p:inverse}

% TODO Outline
%   - Outline simplification of f^{-1}
%   - Closed form for theta, chi
%   - Give algorithm for any w?

\subsection{XOR ($\oplus$) Distributivity} \label{sec:p:xor}

% TODO Outline
%   - Theta, Rho, Pi - free
%   - Chi, Iota - no
%   - Makes analysis harder, which is good

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath,amsthm,amssymb,enumerate,euscript,mathrsfs}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\title{A Collection of Thoughts on the Keccak Construction}

@@

\section{A Series of Introductions} \label{sec:intro}
    This paper presents the work of the author towards the completion of the
requirements for the MATH 490 Independent Study course under Dr. Clifford
Bergman at Iowa State University of Science and Technology. This work was an
extension of the author's Honors Project under Dr. Eric Bergman and utilized
the resulting
\href{https://github.com/cipherboy/hash\_framework}{hash\_framework} project.
Additional artifacts related to this project can be seen in the
\href{https://github.com/cipherboy/keccak-attacks}{keccak-attacks} repository.

@@

    While there have been many applications of mathematics, few are as
demanding and shrouded in as much secrecy as cryptography. Cryptography exists
because of the fundamental need of civilizations, governments, and individuals
to keep secrets secure from devoted adversaries. While modern cryptography
combines the disciplines of mathematics, computer science, and computer
engineering, prior to the turn of the 20th century, cryptography lacked much
of its modern rigor.

    Prior to the 20th century, cryptography was split into three major
branches: ciphers, codes, and stenography.

@@

\subsection{Introduction to Terminology} \label{sec:i:terminology}

Define $\Sigma = \{0, 1\}$ to be the alphabet, and for
$w \in \{1, 2, 4, 8, 16, 32, 64\}$, $S_{w} = \Sigma^{25w}$, to be the set of
all binary strings of length $25w$.

@@

In the following section, we detail various mathematical properties of the five
permutation functions which make up the core round function of Keccak. In most
cases, we seek to provide mathematical proofs of these properties. In all
cases, we rely on external code and Boolean Satisfiability for computerized
proofs of these theorems, in ways independent of the ways proved here. In a
few cases, we provide references into the existing literature.

@@

\subsection{Properties of $\theta$} \label{sec:p:t}

@@

In this section, we show that $\theta$ is bijective, that XOR distributes
through $\theta$, give a method for finding the inverse of $\theta$, and
give the order of $\theta$.

Let $w$ be a fixed power of 2. Since $S_{w}$ is of finite size, it suffices to
show that, $\forall x, y \in S_{w}$,
$x \neq y \Rightarrow \theta(x) \neq \theta(y)$. Assume the hypothesis: then
$x \oplus y \neq 0^{25w}$.

\begin{lemma} \label{lem:p:t:1}
\begin{align*}
    \theta(a) = 0^{25w} \iff a = 0^{25w}
\end{align*}
\end{lemma}
\begin{proof}

This follows from the definition of $\theta$: note that
$\theta(0^{25w}) = 0^{25w}$. If $a \neq 0^{25w}$, then there
exists an index set $I_{1}$ such that $\forall i \in I_{1}$, $a[i] = 1$,
and $\forall j \not\in I_{1}$, $a[j] = 0$. Then $\theta(a) \neq 0^{25w}$,
which follows from the construction of $\theta$. (Note that each output
bit of $\theta$ is composed of the XOR of 11 values in a fixed pattern).

\end{proof}

\begin{lemma} \label{lem:p:t:2}
$\forall a, b \in S_{w}$,
\begin{align*}
    \theta(a \oplus b) = \theta(a) \oplus \theta(b)
\end{align*}
\end{lemma}
\begin{proof}

This follows from the definition of $\theta$: note that $\theta$ is
composed entirely of XORs and that XOR is commutative and associative.

\end{proof}

Combining Lemma \ref{lem:p:t:1} and Lemma \ref{lem:p:t:2}, we have that:

\begin{align*}
    x \oplus y = 0^{25w} & \iff \theta(x \oplus y) = \theta(0^{25w}) \\
        & \iff \theta(x \oplus y) = 0^{25w} \\
        & \iff \theta(x) \oplus \theta(y) = 0^{25w}
\end{align*}

and hence $\theta$ is bijective.

To construct the inverse of $\theta$, note that
$A'[x, y, z] = A[x, y, z] \oplus D[x, z]$; hence,
$A[x, y, z] = A'[x, y, z] \oplus D'[x, z]$ for some $D' = D$. Since $D[x, z]$
is composed of several $C[x, z]$, where
$C[x, z] = \oplus_{y = 0}^{4} A[x, y, z]$, we can similarly define $C'[x, z]$
to be $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$. Then, we expect the inverse
of $\theta$ to be of a similar form. This reduces to a linear algebra problem
over boolean variables. We know that $D'[x, z] = D[x, z]$ in order to recover
$A[x, y, z]$. Hence we can represent $D[x, z]$ as a series of bits of length
$5 \times w$, where bit $i = z' + 5 \times x'$ is $1$ if and only if
$C[x', z']$ is used in the construction of $D[x, z]$. Further, we can view each
of the $C'[x, z]$ as being the conjunction of three $C[x', z']$ in $A$, and thus
can represent these as bit strings where bit $j = z' + 5\times x'$ is 1 if
and only if $C[x', z']$ is used to construct $C'[x, z]$. (That is,
since $C'[x, z] = \oplus_{y = 0}^{4} A'[x, y, z]$,
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus D[x, z])$ and hence
$C'[x, z] = \oplus_{y = 0}^{4} (A[x, y, z] \oplus C[x', z'] \oplus C[x'', z''])$,
and thus $C'[x, z] = C[x, z] \oplus C[x', z'] \oplus C[x'', z'']$, for some
$x, z, x', z', x'', z''$ based on the definition of $\theta$. Thus,
giving each $C'[x, z]$ a constant $c_{x,z}$ for whether it is used in constructing
$D'[x, z]$, we can form a system of linear equations and solve for the
constants $c_{x, z}$ in each expression. Since there are $5 \times w$ variables
and $5 \times w$ equations in each equation for $D'[x, z]$, this can be solved
easily, yielding the inverse of $\theta$.

Lastly, we have computed the order of the permutation $\theta$ for all $w$.
We reproduce them here without proof; they were found by direct search, but
only verified exhaustively for $w = 1$. In general, the order is given by
the expression $3 \times w$.

\begin{tabular}{c c} \label{tab:p:t:1}
    w & Order \\
    1 & 3 \\
    2 & 6 \\
    4 & 12 \\
    8 & 24 \\
    16 & 48 \\
    32 & 96 \\
    64 & 192 \\
\end{tabular}

\subsubsection{Of $\rho$} \label{sec:p:r}

@@

Since $\rho$ is a simple permutation of the location of bits, it holds that
$\rho(a \oplus b) = \rho(a) \oplus \rho(b)$.

It is obvious that the order of the $\rho$ permutation is $w$: this is obvious
from the construction of $\rho$.

\subsubsection{Of $\pi$} \label{sec:p:p}

@@

Since $\pi$ is a simple permutation of the location of bits, it holds trivially
that $\pi(a \oplus b) = \pi(a) \oplus \pi(b)$.

It is obvious that the order of the $\pi$ permutation is $24$: this is obvious
from the construction of $\pi$.

\subsubsection{Of $\chi$} \label{sec:p:c}

@@

It is non-trivial that the order of the $\chi$ function is a constant 4.
However, this can be verified through the use of SAT to prove that $\chi^{4}$
is the identity function. Hence $\chi^{-1} = \chi^{3}$.

\subsubsection{Of $\iota$} \label{sec:p:i}

@@

Note that since $\iota$ is an XOR with a fixed value, it is obvious that
$\iota$ is a bijection: for any $w$, for any $i$, and for all $x \in S_{w}$,
$\iota(\iota(x, i), i) = x$, since $x \oplus \iota_{i} \oplus \iota_{i} = x$.
Hence, $\iota$ is its own inverse and hence $\iota$ is bijective since the
inverse is well defined for all $x \in S_{w}$.

Lastly, note that it is trivial that the order of the $\iota$ permutation is
2 by construction (due to the XOR).

@@
